---
title: "Forecast daily bike rental demand using time series models"
date: "`r Sys.Date()`"
output: html_document
author: "Mateus Camargo de Lima"
---

# About Data Analysis Report

This RMarkdown file contains the report of the data analysis done for the project on forecasting daily bike rental demand using time series models in R. It contains analysis such as data exploration, summary statistics and building the time series models. The final report was completed on `r date()`. 

**Data Description:**
This dataset contains the daily count of rental bike transactions between years 2011 and 2012 in Capital bikeshare system with the corresponding weather and seasonal information.

**Data Source:** 
https://archive.ics.uci.edu/ml/datasets/bike+sharing+dataset

**Relevant Paper:** 
Fanaee-T, Hadi, and Gama, Joao, 'Event labeling combining ensemble detectors and background knowledge', Progress in Artificial Intelligence (2013): pp. 1-15, Springer Berlin Heidelberg



# Task One: Load and explore the data

## Load data and install packages
```{r}

# Install and load required packages
install.packages("timetk")
install.packages("tidyverse")
install.packages("lubridate")
install.packages("forecast")
install.packages("TTR")
install.packages("tseries")

library(timetk)
library(tidyverse)
library(lubridate)
library(forecast)
library(TTR)
library(tseries)

# Load built-in dataset
data("bike_sharing_daily")
bike_data <- bike_sharing_daily
View(bike_data)

# Rebuild date column starting from "2011-01-01"
bike_data <- bike_data %>%
  mutate(date = as.Date("2011-01-01") + days(0:(n() - 1)))
```

## Describe and explore the data
```{r}

# Summary statistics
summary(bike_data)

# Correlation between temperature and rentals
cor(bike_data$temp, bike_data$cnt)
cor(bike_data$atemp, bike_data$cnt)

# Mean and median temperature per season
bike_data %>%
  group_by(season) %>%
  summarise(mean_temp = mean(temp), median_temp = median(temp))

# Monthly averages
bike_data %>%
  mutate(month = month(date)) %>%
  group_by(month) %>%
  summarise(mean_temp = mean(temp), mean_hum = mean(hum), mean_wind = mean(windspeed), total_rentals = mean(cnt))

# Boxplot of temperature by season
boxplot(temp ~ season, data = bike_data, main = "Temperature by Season")

```



# Task Two: Create interactive time series plots
```{r}

# Interactive plot using timetk
bike_data %>%
  plot_time_series(date, cnt, .interactive = TRUE, .plotly_slider = TRUE)

# Seasonal diagnostics
bike_data %>%
  plot_seasonal_diagnostics(date, cnt)

# Anomaly diagnostics
bike_data %>%
  plot_anomaly_diagnostics(date, cnt)

```




# Task Three: Smooth time series data
```{r}

# Convert to time series object
bike_ts <- ts(bike_data$cnt, frequency = 365)

# Clean anomalies and missing values
bike_clean <- tsclean(bike_ts)

# Simple Moving Average (SMA)
bike_sma <- SMA(bike_clean, n = 10)
plot(bike_sma, main = "Simple Moving Average (Order 10)")

# Simple Exponential Smoothing
bike_hw <- HoltWinters(bike_clean, beta = FALSE, gamma = FALSE)
plot(bike_hw)

```



# Task Four: Decompse and access the stationarity of time series data
```{r}

# Decomposition
decomp <- decompose(bike_ts)
plot(decomp)

# Augmented Dickey-Fuller Test for stationarity
adf.test(bike_ts)

# ACF and PACF plots
acf(bike_ts)
pacf(bike_ts)

# Differencing (if necessary)
bike_diff <- diff(bike_ts)
plot(bike_diff)
adf.test(bike_diff)

```



# Task Five: Fit and forecast time series data using ARIMA models
```{r}

# Auto ARIMA model
auto_fit <- auto.arima(bike_clean)
summary(auto_fit)

# Manual ARIMA model example
manual_fit <- arima(bike_clean, order = c(1,1,1))
summary(manual_fit)

# Residual diagnostics
shapiro.test(residuals(auto_fit))
acf(residuals(auto_fit))
pacf(residuals(auto_fit))

# Forecast the next 25 values
forecast_auto <- forecast(auto_fit, h = 25)
plot(forecast_auto, main = "Forecast with Auto ARIMA")

forecast_manual <- forecast(manual_fit, h = 25)
plot(forecast_manual, main = "Forecast with Manual ARIMA")

```



# Task Six: Findings and Conclusions

  - The analysis revealed seasonal and trend patterns in bike rental demand.
  - Temperature and season were found to significantly influence rental counts.
  - ARIMA models (both auto and manual) showed solid predictive performance.
  - Results matched expectations for seasonal time series data.
  - The timetk package and interactive visualizations helped uncover useful insights.
  
  